---
title: "R-Functions"
author: "Data Scientists"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### Data Frames
Yesterday, we have created a data frame from vectors. Today we will load data from built-in datasets in R and work on it

```{r}
library(help="datasets") #You can see the built-in datasets in R
```
Lets load mtcars dataset and work on it

```{r}
data("mtcars")
DF<-mtcars
#to remove variables from the environment we use rm function
rm(mtcars)
```

```{r}
#Lets observe the structure of the data
str(DF)

#Summary of the data
summary(DF)
```
##### Exercise
* Which of the variables of the data are numeric and which of them categorical

#### Subsetting Dataframes
* Subsetting using []
* Using a subset function
```{r}
DF_s1<-DF[c(1,6,7)] #Selecting 1st 6th and 7th colum

x<- c("mpg","wt","qsec")
DF_s2<-DF[names(DF)%in%x]
names(DF_s2)
```
##### Exercise
* Take subset of all those columns that are not there in x
```{r echo=F}
x<- c("mpg","wt","qsec")
DF_s3<-DF[!(names(DF) %in% x)]
names(DF_s3)

#Select all the cars that have mileage over 25 and only mph and cyl columns
DF_s6<-DF[DF$mpg>25,c("mpg","cyl")] #Instead of column names we can also give indices
dim(DF_s6)
```

##### Using subset
* subset(what is to be subset, on what condition, what should be selected)
* Condition is not necessary to select columns but is necessary to select rows
```{r}
DF_s4<-subset(DF,select=c(mpg,cyl)) #Observe that condition is not mentioned and we got the required columns
names(DF_s4)
dim(DF_s4)
##Select all the rows for mpg and cyl whose mpg is greater than 25
DF_s5<-subset(DF,mpg>25,select=c(mpg,cyl))
dim(DF_s5)
```


##### Lets work with the mileage column.. and study conditionals, loops.
* Create a new variable as "condition" that if mileage is less than 25 "Bad" else "Good"
* For this task we use a function ifelse
```{r}
DF$condition<-ifelse(DF$mpg>25,"Good","Bad")
DF$condition
```
##### Ifelse can be nested
* mpg>25 is good, mpg>18 is ok and mpg<18 is bad
```{r}
DF$condition<-ifelse(DF$mpg>25,"Good",ifelse(DF$mpg>18,"OK","Bad"))
DF$condition
```
##### If we had to do this using a loop
```{r}
for(i in 1:nrow(DF)){
  if(DF$mpg[i]>25){
    DF$condition[i]<-"Good"
  }
  else{
    DF$condition[i]<-"Bad"
  }
}
```

```{r}
for(i in 1:nrow(DF)){
  if(DF$mpg[i]>25){
    DF$condition[i]<-"Good"
  }
  else 
    if(DF$mpg[i]>18){
    DF$condition[i]<-"OK"
  }
    else{
    DF$condition[i]<-"Bad"
  }
  
}

DF$condition
```

#### Writing a custom function in R
* Lets create a simple functions to check if a number is even or odd
```{r}
even_odd<-function(x){
if(x%%2==0){
return("even")
}
else{
return("odd")
}

}

even_odd(20)
```

#### While loop
```{r}
i = 1
while (i < 6){
  print(i)
  i = i+1
}
```

#### next and break statements
```{r}
x = 1:5

for (val in x) {
  if (val == 3){
    break
  }
  print(val)
}

# next statement

for (val in x) {
  if (val == 3){
    next
  }
  print(val)
}

```


##### Exercise
*Create a function to obtain the sum of elements of a vector. The argument that the function takes is vector
* Create a function that takes two arguments a,b and returns a raised to power b

##### Extra credit
* Create a function to return min and max value from a vector
```{r echo=F}
additon<-function(x){
  return(sum(x))
}

A=1:10
additon(A)

pow<-function(x,y){
  return(x^y)
}


min_max1<-function(x){
  
  #return(list(c(min(x),max(x))))
  return(c(min(x),max(x)))
}
A=1:10
min_max1(A)


```

```{r}
# Multiple Returns
# The return() function can return only a single object. If we want to return multiple values in R, we can use a list (or other objects) and return it.
multiReturn = function() {
  myList <- list(1, 20, c("a","b"))
  return(myList) 
}

multiReturn()
```

#### Lists
* A list is datastructure that can store multiple data structures
```{r}
val=c(1,7,9)
ch=c("A","X","Z")
A<-list(val,ch,data.frame(val,ch),list(val,ch))
```

To access the elements of the list
```{r}
A[1] #To get the first element of the list which is val. But we know that val has 3 elements in it
A[[1]][1] # To get the first element of the val

A[[3]]$ch #Since the third element of the list is a dataframe we can access the elements of the dataframe using $

##elements of list can also be accessed using names if they have
B=list(M=val,N=ch,O=data.frame(val,ch),P=list(val,ch))
B$O[[1]]
```
To unlist the elements
```{r}
unlst<-unlist(B)
unlst
```

#### Subsetting a data frame
* In the morning we have learnt about numeric and categorical attributes. Lets separate out numeric attributes and work on them
  + By indices
  + By column names
```{r}
names(DF)
DF_numeric<-DF[c(1,3:6)]
```

### The "Apply" family
* Used to manipulate slices of data in a repetitive way

##### Apply
* Syntax apply(on what, by row or by column, what function to be applied)
```{r}
#Lets say in the num_DF, we need to find mean value for each of the columns. This is a repetitive process that can be quickly done by apply
apply(DF_numeric,2,mean)

#Although it is not meaningful in this case, if we want to take mean of each row
apply(DF_numeric,1,mean)
```
We can also apply custom functions on the data. Lets say we want to take log of x for each value
```{r}
flog<-function(x){
  return(log(x))
}
apply(DF_numeric,2,flog)
```

##### Excerise
* Write a function that takes a vector to identify how far a value is away from its mean. Use apply function to apply it on DF_numeric
```{r echo=FALSE}
f_dist<-function(x){
  return(x-mean(x))
}

apply(DF_numeric,2,f_dist)

```

##### tapply
* We have a data thaqt can be divided into groups
* On each of this group we want to apply the function and want to get a table as output.
* We use tapply

Lets us consider mpg and cyl of DF data, we want to find the mean mileage for each of the group.
```{r}
#Observe the data type of mpg and cyl
str(DF$mpg)
str(DF$cyl)
#Since cyl is interpreted as numeric lets covert that into categorical
DF$cyl<-as.factor(as.character(DF$cyl)) #Observe this syntax 
str(DF$cyl) #Observe that the data type is converted to factor i.e. categorical

#Now for each of the cyl group lets compute the mean
tapply(DF$mpg,DF$cyl,FUN=mean)
```
##### sapply and lapply
* Both work in a similar way, traverse over a vector or list and calling the function for each item except that sapply simplifies the output 
```{r}
sapply(DF_numeric,mean)
lapply(DF_numeric,mean)
sapply(DF_numeric,mean,simplify = F)
```

#### Working with Factors/Categories
* Factors have levels
* Factors created when we read non-numerical columns into a data frame.
```{r}
x <- factor(c("A","B","C","A"))
str(x) #Observe that individual elements are stored as numbers

```
To get the levels of the variable
```{r}
levels(x)
```
###### Here is an exercise
* Create a numeric variable A with 5 values in it
* Create a categorical variable B with 4 levels in it "A","B","C" and 5 elements
* Replace the 5th element of variable A with 1000
* Replace the 5th element of Variable B with "Alpha"
```{r}
A<-c(1,2,3,4,5)
B<-factor(c("A","A","B","C","B"))
A
B
A[5]<-1000
A
B[5]<-"Alpha"
B
```

##### How to modify a factor
```{r}
levels(B)<-c(levels(B),"Alpha")
B[5]<-"Alpha"
B
```

##### Exercise
* Create a factor vector with levels "A","B","C" with 20 elements
* All the elements with level B are to be converted to "Beta"
```{r}
X<-factor(sample(c("A","B","C"),20,replace = T))
X
levels(X)[levels(X)=="B"]<-"Beta"
X

```

